# Data wrangling for model assessment

```{r, echo = FALSE, message=FALSE, eval=TRUE}
library(knitr)
opts_chunk$set(eval = FALSE, message = FALSE, error = TRUE)
library(tweetrmd)
```

We'll talk about the `dplyr` package, and assessing models. 

## dplyr verbs

We have already seen some `dplyr` verbs, but we are going to need more as we move on. So, let's spend some time focusing on them. Here are the main verbs we will cover:

verb | action | example
-------|-------------|-------
`select()` | take a subset of *columns* | `select(x,y)`, `select(-x)`
`filter()` | take a subset of *rows* | `filter(x == __, y > __)`
`arange()` | reorder the *rows* | `arrange(x)`, `arrange(desc(x))`
`summarize()` | a many-to-one or many-to-few summary | `summarize(mean(x), median(y))` 
`mutate()` | a many-to-many operation that creates a new variable | `mutate(x = ___, y = ___)`
`group_by()` | group the *rows* by a specified *column* | `group_by(x) %>% something()`


We'll use the `babynames` package to play with the verbs, so let's begin by loading that dataset.

```{r}
library(babynames)
babynames
```

We could `skim` the data, to learn something about it:

```{r}
library(skim)
skim(babynames)
```

### select()

```{r}
select(babynames, name, prop)
```

1. Alter the code to select just the `n` column

#### select() helpers

- `:` select range of columns, `select(storms, storm:pressure)`
- `-` select every column but `select(storms, -c(storm, pressure))`
- `starts_with()` select columns that start with... `select(storms, starts_with("w"))`
- `ends_with()` select columns that end with... `select(storms, ends_with("e"))`
- ...and more! Check out the Data Transformation cheatsheet

### filter()
extract rows that meet logical criteria 

```{r}
filter(babynames, name == "Amelia")
```

Notice I'm using `==`, which tests if things are equal. In R, `=` sets something. There are other logical comparisons you can use

| | |
|------|------------|
x < y | less than
x > y | greater than 
x == y | equal to
x <= y | less than or equal to 
x >= y | greater than or equal to
x != y | not equal to
x %in% y | group membership
is.na(x) | is NA
!is.na(x) | is not NA

1. Now, see if you can use the logical operators to manipulate our code to show:

-	All of the names where `prop` is greater than or equal to 0.08
- All of the children named "Sea"
- All of the names that have a missing value for `n`
(Hint: this should return an empty data set).

Common mistakes:

- using `=` instead of `==`
- forgetting quotes

We can also filter rows that match *every* logical criteria,

```{r}
filter(babynames, name == "Amelia", year == 1880)
```

For this, you need to use Boolean operators

| | |
---|----|
a & b | and 
a \| b | or
xor(a, b) | exactly or
!a | not
a %in% c(a, b) | one of (in)

2. Use Boolean operators to alter the code below to return only the rows that contain:

- Girls named Sea
- Names that were used by exactly 5 or 6 children in 1880
- Names that are one of Acura, Lexus, or Yugo

### arrange()

Orders rows from smallest to largest values


```{r}
arrange(babynames, n)
```

1. Arrange babynames by n. Add prop as a second (tie breaking) variable to arrange by.
2. Can you tell what the smallest value of n is? Any guesses why?

Another helpful function is `desc()`, which changes the ordering to largest smallest,

```{r}
arrange(babynames, desc(n))
```

3. Use `desc()` to find the names with the highest `prop`.
4. Use `desc()` to find the names with the highest `n`.

## %>%, the pipe



## Modeling conditions

### Least squares

When `R` finds the line of best fit, it is minimizing the sum of the squared residuals, 

$$
			SSE = \sum_{i=1}^n (y_i - \hat{y_i})^2, \qquad \hat{\sigma}_\epsilon = \sqrt{\frac{SSE}{n-2}}
$$
in order for the model to be appropriate, a number of conditions must be met. 

- **L**inearity
- **I**ndependence
- **N**ormality
- **E**quality of variance

These conditions are mainly related to the distribution of the residuals. 



Much of this code was adapted from [Master the Tidyverse](https://github.com/rstudio-education/master-the-tidyverse)
